### 2020-10-01 doc/manuscript/Dow_manuscript.Rmd
intro edits

As global atmospheric greenhouse gas levels are rising, and little meaningful policy is being implemented to combat this, our planet is expected to surpass the 2.5 degree C warming mark that the Paris accord pledged to avoid. The absence of human intervention increases our reliance on natural systems to soften the blow of climate change. Plant life, especially forests, make up a large portion of these natural systems, offering many valuable ecosystem services. For one, they house an astonishing 80% of terrestrial species and are capable of removing the greenhouse gas CO2 from the atmosphere and storing it in the trees or soil. Forests sequester about one-third of all anthropogenic CO2 and store ~45% of all terrestrial carbon (Bonan, 2008). However, their future under the current era of rapid change is uncertain (Friedlingstein et al, 2006).

@@

In recent decades, an increase in average Spring temperatures has been attributed to climate change ().  The effects of this Spring warming on leaf phenology have been extensively studied (Way, 2011; ) leading to many interesting observations. As temperatures are increasing, some regions have seen earlier leaf out in deciduous trees, sometimes matched or exceeded by earlier leaf senecence, meaning a potential shift in growing season timing (Ibáñez et al., 2010; Jeong et al., 2011). The timing of Spring warming also has a significant effect on leaf phenology response. Warming events later in Spring were shown to impact leaf emergence timing more than earlier Spring warming (Freidl et al, 2014) and earlier leaf emergence has been shown to cause complex lagged effects on subsequent seasonal growth in European broad leaf deciduous and mixed forests (Crabbe et al, 2016).

However, the fine-scale responses of temperate deciduous species to a warming Spring are not fully understood (source?). Many observations of phenological shifts like these are based on GIS, eddy-covarience measurements, or other broad scale measurements of phenological timing, and do little to discern differences between individual species, where leaf and stem-growth phenology can be decoupled (D'Orangeville et al, in review). Shifts in leaf phenology timing could be accompanied by a change in stem-growth phenology, which would likely affect woody productivity, carbon and nutrient cycling, and water use patterns in temperate forests. In reality, we know little about the stem-growth phenology of temperate deciduous species — let alone how these are affected by an early spring.

To date, very little research, has been done on the phenology of stem-growth. Timing of intraannual growth is extremely important as it can dictate periods of highest drought sensitivity of NPP in temperate forests (Foster, 2014; ). For example, if a drought were to occur when trees were at their peak growth rates, it could decrease that year's NPP substantially, and subsequently, reduce the total amount of carbon sequestered creating a negative climate feedback loop (source?). Recent results from D’Orangeville et al. (in review) have shown that wood-type plays a significant role in the timing of leaf and wood phenology. They showed that ring-porous trees begin growth earlier in the season compared to diffuse-porous trees, but both reached 75% total annual growth around the same time. This implies evolutionary differences in the mechanisms controlling growth of these species, which have the potential to expose each group to different climate stresses (source for RP vs DP?).

While not entirely impossible, it is difficult to record intraannual growth, or compare interspecific responses to Spring warming using leaf emergence timing alone. A solution to this is the use of dendrometer bands. Dendrometer bands allow a look at individual trees’ intraannual growth at a fine temporal scale (source?). Using many measurements taken within a year, we can model the growth of individual trees and extract growth milestones and variables for interspecific, intraspecific, and temporal comparison (McMahon & Parker, 2014). This allows the determination of variable responses of all tree species within a single forest, meaning a better understanding of future growth under projected climate conditions. Here, we use 9 years of biweekly dendrometer band measurements for 4 species to characterize intraannual stem-growth phenology within the ForestGEO plot at SCBI, calculating for each tree in each year where data was available: the day of year (DOY) where 25, 50, and 75% annual growth was achieved; the maximum growth rate; the DOY where maximum growth rate was achieved; and the total growth. Correlation between these variables and SPring temperatures over the 9 year study period (2011-2019) were assessed using a heirarchal bayesian model.

>>>>

As global atmospheric greenhouse gas levels are rising, and little meaningful policy is being implemented in response, our planet is expected to surpass the 2.5 degree C warming mark that the Paris accord pledged to avoid. The absence of human intervention increases our reliance on natural systems to soften the blow of climate change. Plant life, especially forests, make up a large portion of these natural systems, offering many valuable ecosystem services. For one, they house an astonishing 80% of terrestrial species and are capable of removing the greenhouse gas CO2 from the atmosphere and storing it in the trees or soil. Forests sequester about one-third of all anthropogenic CO2 and store ~45% of all terrestrial carbon (Bonan, 2008). However, their future under the current era of rapid change is uncertain (Friedlingstein et al, 2006).

@@

In recent decades, an increase in average Spring temperatures has been attributed to climate change. This relatively small increase has already been impacting many natural systems including forests (Parmesan & Yohe, 2003; Zhihong, 2006). The effects of this Spring warming on leaf phenology have been extensively studied (Way, 2011; Klosterman & Hufkins & Richardson, 2018) leading to many interesting observations. As temperatures are increasing, some regions have seen earlier leaf out in deciduous trees, sometimes matched or exceeded by earlier leaf senescence, meaning a potential shift in growing season timing or length (Ibáñez et al, 2010; Jeong et al., 2011). The timing of Spring warming also has a significant effect on leaf phenology response. Warming events later in Spring were shown to impact leaf emergence timing more than earlier Spring warming (Freidl et al, 2014) and earlier leaf emergence has been shown to cause complex lagged effects on subsequent seasonal growth in European broad leaf deciduous and mixed forests (Crabbe et al, 2016). Another study found that the velocity of leaf development is correlated to bud burst timing, such that the later in the Spring bud burst occurred, the faster leaves developed and vice versa (Klosterman & Hufkins & Richardson, 2018).

However, the fine-scale responses of temperate deciduous species to a warming Spring are not commonly researched due to the increased effort required. Shifts in leaf phenology timing could be accompanied by a change in stem-growth phenology, which would likely affect woody productivity, carbon and nutrient cycling, and water use patterns in temperate forests (Keeling et al, 1996;). Many observations of climate induced phenological shifts are based on GIS, eddy-covarience measurements, or other broad scale measurements of phenological timing, and do little to discern differences between individual species, where leaf and stem-growth phenology can be decoupled (D'Orangeville et al, in review). To date, very little research, has been done on the phenology of stem-growth.

In fact, we know little about the stem-growth phenology of temperate deciduous species — let alone how these are affected by changes in Spring temperature patterns. Timing of intraannual stem-growth is extremely important as it could have profound impacts on ecosystem interactions, as well as dictate periods of highest drought sensitivity of NPP in temperate forests (Foster, 2014; ). For example, if a drought were to occur when trees were at their peak growth rates, it could decrease that year's NPP substantially, and subsequently, reduce the total amount of carbon sequestered contributing to a negative climate feedback loop (source?). Recent results from a paper by D’Orangeville et al. (in review) have shown that wood-type plays a significant role in the timing of leaf and wood phenology. They showed that ring-porous trees begin growth earlier in the season compared to diffuse-porous trees, but both reached 75% total annual growth around the same time. This implies evolutionary differences in the mechanisms controlling growth of these species, which have the potential to expose each group to different climate stresses (source for RP vs DP?).

While not entirely impossible, it is difficult to record intraannual growth or compare interspecific responses to Spring warming using leaf-phenology or other indirect observations. A solution to this is the use of dendrometer bands. Dendrometer bands allow a look at individual trees’ intraannual growth at a fine temporal scale (source?). Using many measurements taken within a year, we can model the growth of individual trees and extract growth milestones and variables for interspecific, intraspecific, and temporal comparison (McMahon & Parker, 2014). This allows the determination of variable responses of all tree species within a single forest, meaning a better understanding of future growth under projected climate conditions. Here, we use 9 years of biweekly dendrometer band measurements for 4 species to characterize intraannual stem-growth phenology within the ForestGEO plot at SCBI, calculating for each tree in each year where data was available: the day of year (DOY) where 25, 50, and 75% annual growth was achieved; the maximum growth rate; the DOY where maximum growth rate was achieved; and the total growth. Correlation between these variables and Spring temperatures over the 9 year study period (2011-2019) were assessed using a linear mixed model in a heriarchal bayesian framework.
